### 2020-10-26 latex/vlp.tex
Made all edits suggested by Manfredi on Oct 19
Added references to P-B and stable manifold
Also edited more of the VLP proof, it's not quite done but it's coming
along nicely.

The total mechanical energy of the system restricted to the constraint manifold
is given by (\ref{eqn:vlp-energy}), where \(L(q,p)\) and \(\dot{L}(q,p)\) are
known.
\begin{equation}\label{eqn:vlp-energy}
   E(q,p) = \frac{1}{2} \frac{p^2}{mL^2} + \frac{1}{2}\dot{L}^2 - mgL\cos(q)
\end{equation}
In the rest of this chapter, we will derive a \(C^2\) function \(L(q,p)\) based
on natural human motion.
This function will produce constrained dynamics that inject energy into the VLP.

@@

\begin{thm}\label{thm:vlp-energy-stabilization}
   Define \(\theta := \arctan_2(p,q)\).
   Let \(L : \mathbb{S}^1 \rightarrow [\underline{l},\overline{l}]\) be a
   \(C^2\) function of \(\theta\).
   A regular VNHC of the form 
   \(h(\mathbf{q},\mathbf{p}) = l - L(\theta)\) injects energy into the VLP
   if there exists 
   \(l_\text{avg} \in [\underline{l},\overline{l}]\) such that 
   \begin{equation}\label{eqn:vlp-energy-gain-condition}
      \left(l_\text{avg} - L(\theta)\right)sin(2\theta) \geq 0 
      \text{ }\forall \theta \in \mathbb{S}^1
      ,
   \end{equation}
   with the property that the inequality is strict except at the
   coordinate axes.
   If the inequality is flipped, the VNHC dissipates energy. 
\end{thm}
\begin{proof}
   The constraint manifold for \(h\) is the cylinder \(\Gamma = \mathbb{S}^1 \times \R \). 
   In negative-time, the dynamics on \(\Gamma\) are
   \begin{equation}\label{eqn:negative-time-vlp}
      f(q,p) = 
      -\begin{bmatrix}
         \dot{q} \\ \dot{p}
      \end{bmatrix} = \begin{bmatrix}
         -\frac{p}{mL(\theta)^2} \\
         mgL(\theta)\sin(q)
      \end{bmatrix}
      .
   \end{equation}
   This has equilibria at \((q,p) = (0,0)\) and \((q,p) = (\pi,0)\) (the
   third ``equilibrium" \((-\pi,0)\) is the same as \((\pi,0)\) because 
   \(q \in \mathbb{S}^1\)).
   To prove the origin of the positive-time system (\ref{eqn:vlp-hamiltonian})
   is a repeller, we will show the origin of the negative-time system 
   (\ref{eqn:negative-time-vlp}) is almost everywhere globally asymptotically
   stable.

@@

   The linearization of (\ref{eqn:negative-time-vlp}) at the upright equilibrium
   \((\pi, 0)\) is given by
   \[
      df_{(\pi,0)} = \begin{bmatrix}
         0 & -\frac{1}{mL(0)^2} \\
         -m g L(0) & 0 \\
      \end{bmatrix}
      ,
   \]
   which has eigenvalues at \(\lambda = \pm \sqrt{\frac{g}{L(0)}}\).
   The positive-time system (\ref{eqn:vlp-hamiltonian}) has the same eigenvalues
   at this equilibrium.
   Label the set of initial conditions converging to \((\pi, 0)\) in
   positive time by
   \[
      \Pi_0^+ := \left\{ (q_0,p_0) \in \Gamma \mid \lim_{t \to \infty} (q(t),p(t)) = (\pi, 0) \right\}
      ,
   \]
   and the set of initial conditions converging to \((\pi,0)\)
   in negative time by \(\Pi_0^-\).
   Since only one eigenvalue is negative, the stable manifold theorem tells us
   that both \(\Pi_0^+\) and \(\Pi_0^-\) are one-dimensional.
   While this implies the origin is not globally asymptotically stable for
   (\ref{eqn:negative-time-vlp}), we can still show it is asymptotically stable
   for almost every initial condition because \(\Pi_0^+\) and \(\Pi_0^-\) are of
   measure zero within \(\Gamma\).

   Define \(D := \Gamma \backslash \Pi_0^-\) to be the domain of interest.
   Using a Krasovskii-LaSalle argument \cite{krasovskii_lasalle}, 
   we will prove that (\ref{eqn:negative-time-vlp}) asymptotically converges to
   \((0,0)\) if it is initialized on \(D\).
   First, we define the zero-set
   \[
      Z = \left\{(q,p) \in \Gamma \mid 
      \dot{E}_\text{avg}(q,p) = 0 \right\}
      .
   \]
   It is easy to show that \(Z\) is the union of the coordinate axes
   \[
      Z= \mathbb{S}^1 \times \left\{0\right\} \cup 
         \left\{0\right\} \times \R
      .
   \]
   Krasovskii-LaSalle tells us that solutions to (\ref{eqn:negative-time-vlp})
   converge to the largest invariant set in \(Z \cap D\).
   Since \(f(q,p) = (0,0)\) on \(D\) only when \((q,p) = (0,0)\),
   the origin is asymptotically stable on \(D\) for the negative-time system.
   Hence, the positive-time system is diverging away from the origin 
   in the \((q,p)\) plane, which means the origin is a repeller.

   We now prove the system escapes compact sets in finite time.
   Let \(x(t) = (q(t),p(t))\) be the orbit for the positive-time system.
   Let \(K \subset \Gamma\) be compact and assume \(x(0) \in K\backslash \Pi_0^+\).
   For a proof by contradiction, suppose that for all \(T > 0\) there exists
   \(t' > T\) where \(x(t') \in K\).
   Let 
   \[
      k := \max\limits_{(q,p) \in K} E_\text{avg}(q,p)
      ,
   \] 
   be the largest energy value attained in \(K\), and let its sublevel set be
   \[
      E_k := \left\{ (q,p) \in \Gamma \mid E_\text{avg}(q,p) \leq k \right\}
      .
   \]
   All sublevel sets of \(E_\text{avg}\) are compact, negatively
   invariant (by (\ref{eqn:vlp-proof-neg-inv})), and \(E_k\) contains \(K\).
   We assumed \(x(t') \in K\), so it follows that \(x(t') \in E_k\). 
   Then, since \(E_k\) is negatively invariant, the half orbit
   \(x([0,t'])\) is contained fully within \(E_k\).
   In particular, \(x(T) \in E_k\);
   but \(T\) was chosen arbitrarily, which implies \(x(t) \in E_k\) for all
   \(t \in [0,\infty[\).
   Since \(E_k\) is compact, \(x(t)\) has a positive limit set in
   \(E_k\) which (by Poincar\'{e}-Bendixson) is either an equilibrium or a
   closed orbit \textbf{SOURCE?}.
   The system cannot have closed orbits, since this would violate asymptotic
   stability of \((0,0)\) in negative time. 
   Hence, the positive limit set must be one of the equilibria. 
   The origin is a repeller, which means \(x(t)\) must be converging to 
   \((\pi,0)\).
   This is a contradiction, since we assumed \(x(0) \not \in \Pi_0^+\).

>>>>

The total mechanical energy of the system restricted to the constraint manifold
is given by (\ref{eqn:vlp-energy}), where \(L(q,p)\) and \(\dot{L}(q,p)\) are
known.
\begin{equation}\label{eqn:vlp-energy}
   E(q,p) = \frac{1}{2} \frac{p^2}{mL^2} + \frac{1}{2}\dot{L}^2 - mgL\cos(q)
   .
\end{equation}
In the rest of this chapter, we will derive a \(C^2\) function \(L(q,p)\) based
on natural human motion.
This function will produce constrained dynamics that inject energy into the VLP.

@@

\begin{thm}\label{thm:vlp-energy-stabilization}
   Define \(\theta := \arctan_2(p,q)\).
   Let \(L : \mathbb{S}^1 \rightarrow [\underline{l},\overline{l}]\) be a
   \(C^2\) function of \(\theta\).
   A regular VNHC of the form 
   \(h(\mathbf{q},\mathbf{p}) = l - L(\theta)\) injects energy into the VLP
   if there exists 
   \(l_\text{avg} \in [\underline{l},\overline{l}]\) such that 
   \begin{equation}\label{eqn:vlp-energy-gain-condition}
      \left(l_\text{avg} - L(\theta)\right)sin(2\theta) \geq 0 
      \text{ }\forall \theta \in \mathbb{S}^1
      ,
   \end{equation}
   with the property that the inequality is strict except at the
   coordinate axes.
   If the inequality is flipped, the VNHC dissipates energy. 
\end{thm}
\begin{proof}
   The constraint manifold for \(h\) is the cylinder \(\Gamma = \mathbb{S}^1 \times \R \). 
   In negative-time, the dynamics on \(\Gamma\) are
   \begin{equation}\label{eqn:negative-time-vlp}
      f(q,p) = 
      -\begin{bmatrix}
         \dot{q} \\ \dot{p}
      \end{bmatrix} = \begin{bmatrix}
         -\frac{p}{mL(\theta)^2} \\
         mgL(\theta)\sin(q)
      \end{bmatrix}
      .
   \end{equation}
   Because \(q \in \mathbb{S}^1\), the only equilibria of these dynamics are 
   at \((q,p) = (0,0)\) and \((q,p) = (\pi,0)\).
   To prove the origin of the positive-time system (\ref{eqn:vlp-hamiltonian})
   is a repeller, we will show the origin of the negative-time system 
   (\ref{eqn:negative-time-vlp}) is almost everywhere globally asymptotically
   stable.

@@

   The linearization of (\ref{eqn:negative-time-vlp}) at the upright equilibrium
   \((\pi, 0)\) is given by
   \begin{equation}\label{eqn:linearized-vlp-neg-time}
      df_{(\pi,0)} = \begin{bmatrix}
         0 & -\frac{1}{mL(0)^2} \\
         -m g L(0) & 0 \\
      \end{bmatrix}
      ,
   \end{equation}
   which has eigenvalues at \(\lambda = \pm \sqrt{\frac{g}{L(0)}}\).
   The positive-time system (\ref{eqn:vlp-hamiltonian}) has the same eigenvalues
   at this equilibrium.
   Label the set of initial conditions converging to \((\pi, 0)\) in
   positive time by
   \[
      \Pi_0^+ := \left\{ (q_0,p_0) \in \Gamma \mid \lim_{t \to \infty} (q(t),p(t)) = (\pi, 0) \right\}
      ,
   \]
   and the set of initial conditions converging to \((\pi,0)\)
   in negative time by \(\Pi_0^-\).
   Taking the sublevel set
   \[
      E_{(\pi,0)} := \left\{ (q,p) \in \Gamma \mid E_\text{avg}(q,p) <
         E_\text{avg}(\pi,0) \right\}
      ,
   \] 
   it is clear form (\ref{eqn:vlp-proof-neg-inv}) that 
   \(\Pi_0^+ \subset E_{(\pi,0)}\)
   and 
   \(\Pi_0^- \subset \left(\Gamma \backslash E_{(\pi,0)} \right) \cup (\pi,0)\).
   Since only one eigenvalue of (\ref{eqn:linearized-vlp-neg-time})
   is negative, the stable manifold theorem tells us that both \(\Pi_0^+\) and
   \(\Pi_0^-\) are one-dimensional \cite{stable_manifold}.
   This implies the origin cannot be globally asymptotically stable for
   (\ref{eqn:negative-time-vlp}); however, we can still show it is asymptotically stable
   for almost every initial condition because \(\Pi_0^+\) and \(\Pi_0^-\) are of
   measure zero within \(\Gamma\).

   Define \(D := \Gamma \backslash \Pi_0^-\) to be the domain of interest.
   Using a Krasovskii-LaSalle argument \cite{krasovskii_lasalle}, 
   we will prove that (\ref{eqn:negative-time-vlp}) asymptotically converges to
   \((0,0)\) if it is initialized on \(D\).
   First, we define the zero-set
   \[
      Z = \left\{(q,p) \in \Gamma \mid 
      \dot{E}_\text{avg}(q,p) = 0 \right\}
      .
   \]
   It is easy to show that \(Z\) is the union of the coordinate axes
   \[
      Z= \mathbb{S}^1 \times \left\{0\right\} \cup 
         \left\{0\right\} \times \R
      .
   \]
   Krasovskii-LaSalle tells us that solutions to (\ref{eqn:negative-time-vlp})
   converge to the largest invariant set in \(Z \cap D\).
   Since \(f(q,p) = (0,0)\) on \(D\) only when \((q,p) = (0,0)\),
   the origin is asymptotically stable on \(D\) for the negative-time system.
   In particular, the origin is asymptotically stable for every initial
   condition in the interior of \(E_{(\pi,0)}\).
   Hence, the positive-time system is diverging away from the origin 
   in the \((q,p)\) plane, which means the origin is a repeller.

   We now prove the system escapes compact sets in finite time.
   Let \(x(t) = (q(t),p(t))\) be the orbit for the positive-time system.
   Let \(K \subset \Gamma\) be compact and assume \(x(0) \in K\backslash \Pi_0^+\).
   For a proof by contradiction, suppose that for all \(T > 0\) there exists
   \(t' > T\) where \(x(t') \in K\).
   Let 
   \[
      k := \max\limits_{(q,p) \in K} E_\text{avg}(q,p)
      ,
   \] 
   be the largest energy value attained in \(K\), and let \(E_k\) be its
   sublevel set.
   All sublevel sets of \(E_\text{avg}\) are compact, negatively
   invariant (by (\ref{eqn:vlp-proof-neg-inv})), and \(E_k\) contains \(K\).
   We assumed \(x(t') \in K\), so it follows that \(x(t') \in E_k\). 
   Then, since \(E_k\) is negatively invariant, the half orbit
   \(x([0,t'])\) is contained fully within \(E_k\).
   In particular, \(x(T) \in E_k\);
   but \(T\) was chosen arbitrarily, which implies \(x(t) \in E_k\) for all
   \(t \in [0,\infty[\).
   Since \(E_k\) is compact, \(x(t)\) has a positive limit set in
   \(E_k\) which (by Poincar\'{e}-Bendixson) is either an equilibrium or a
   closed orbit \cite{poincare_bendixson}.
   The system cannot have closed orbits, since this would violate asymptotic
   stability of \((0,0)\) in negative time. 
   Hence, the positive limit set must be one of the equilibria. 
   The origin is a repeller, which means \(x(t)\) must be converging to 
   \((\pi,0)\).
   This is a contradiction, since we assumed \(x(0) \not \in \Pi_0^+\).
